% The document will be printed one side in A4 paper.
%\documentclass[12pt,a4paper,oneside]{report}
%\documentclass[12pt,a4paper,twoside]{report}
\documentclass[12pt,a4paper,twoside]{article}

\usepackage[round]{natbib}

\usepackage[pdftex]{color,graphicx}
\usepackage{setspace}
\usepackage{a4wide}

\usepackage[a4paper]{geometry}
\usepackage[cm]{fullpage}


%subfigures deprecated
%\usepackage{subfigure}
%use subcaption instead
\usepackage{caption}
\usepackage{subcaption}
%side caption: SCfigure
%\usepackage{sidecap}
% Use PDF output.
% The output should be wide.
\usepackage{url}
%for definition list
\usepackage{enumitem}
%for celsius
%\usepackage{gensymb}
%for listing code
\usepackage{listings}
%for table cell with diagonal division
\usepackage{slashbox}
\usepackage{tabularx}
\usepackage{amsmath}

%\usepackage[graphics,tightpage,active]{preview}
%\PreviewEnvironment{tikzpicture}
%\PreviewEnvironment{equation}
%\PreviewEnvironment{equation*}
%\newlength{\imagewidth}
%\newlength{\imagescale}
%\pagestyle{empty}
%\thispagestyle{empty}
%\usepackage{standalone}

\usepackage[french, greek, english]{babel}
%input encoding
%\usepackage[iso-8859-7]{inputenc}
%\usepackage[latin1]{inputenc}
%output encoding
\usepackage[T1]{fontenc}
\selectlanguage{english}

% Pour la dedicace
\usepackage{frcursive}

\usepackage[table]{xcolor}
\usepackage{booktabs}
\usepackage{tabu}

\usepackage[section,subsection,subsubsection]{extraplaceins}
\usepackage{url} %tablerules
\usepackage{xr}
\externaldocument{supplementary}

\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
\usepackage{authblk}


\setcounter{secnumdepth}{0}



%\doublespacing

%\title{Leveraging qPCR with SNP data achieves large scale case-control analysis of KIR3DL1/3DS1 copy number variation in type 1 diabetes}
\title{Predicting \emph{KIR3DL1/S1} copy number from ImmunoChip signals in large scale type 1 diabetes association study}
%and reveals no association in type 1 diabetes}
\author[1]{Nikolas Pontikos}
\author[1]{Deborah Smyth}
\author[1]{Helen Schuilenburg}
\author[1]{Joanna MM Howson}
\author[1]{Neil M. Walker}
\author[2]{Jyothi Jayaraman}
\author[2]{Wei Jiang}
\author[2]{James Traherne}
\author[2]{John Trowsdale}
\author[1]{John A. Todd}
\author[1]{Chris Wallace}
\affil[1]{JDRF/Wellcome Trust Diabetes and Inflammation Laboratory, Cambridge Institute for Medical Research, University of Cambridge, Wellcome Trust/MRC Building, Cambridge CB2 0XY, United Kingdom;}
\affil[2]{Division of Immunology, Department of Pathology, University of Cambridge, Cambridge CB2 1QP, United Kingdom}

\renewcommand\Authands{ and }


\date{}


\begin{document}

\newlength{\oparindent}
\setlength{\oparindent}{\parindent}
\setlength{\parindent}{0pt}

\maketitle
\noindent


%\newpage
%\thispagestyle{empty} 
%\addtocounter{page}{-1}
%\vspace*{\fill}
%\begin{center}
%{\fontfamily{frc}\selectfont \foreignlanguage{french} {Pour ma petite Maman ch\'erie, \`a qui je dois tout \ldots}}
%{\fontfamily{frc}\selectfont Pour ma petite Maman cherie, a qui je dois tout \ldots}
%\end{center}
%\vspace*{\fill}
%\clearpage

\abstract{
Killer Immunoglobulin-like Receptors (KIR) are important surface receptors of Natural Killer cells.
KIRs mediate the fate of target cells based on the composite inhibiting/activating signal generated on binding to their corresponding HLA (Human Leukocyte Antigen) class I ligands.
Type 1 diabetes (T1D) is an autoimmune disease strongly associated with genetic variation in the HLA region, primarily with Class II genes, but also with HLA Class I loci including HLA-Bw4, an epitope grouping both HLA-A and HLA-B alleles into HLA-Bw4-80I and HLA-Bw4-80T subgroups.
Amongst the 17 known genes of the KIR region, characterised by alternative haplotypes of variable gene copy numbers, \emph{KIR3DL1} is the only one proven to bind in-vitro with both HLA-Bw4 epitopes.
Its activating counterpart, \emph{KIR3DS1}, putatively binds the HLA-Bw4-80I epitope only.
\emph{KIR3DL1/S1} is therefore an interesting T1D candidate gene but since the KIR region is poorly characterised, genotyping with conventional SNP arrays is unreliable.
So far, insufficiently powered PCR-based genotyping studies of less than 600 individuals, have failed to detect convincing association of T1D with presence or absence of \emph{KIR3DL1/S1}.  
For the first time, we are able to test for association in a sample size of 6,744 cases and 5,362 controls, twenty-fold larger than any previous study, by using \emph{KIR3DL1/S1} copy number calls from a recently developed KIR qPCR assay in order to impute copy number in samples assayed only with ImmunoChip,
a custom Illumina genotyping chip with probes for 30 SNPs in the target region.
%with dense coverage of autoimmune regions

%The \emph{KIR3DL1/S1} copy numbers from qPCR data were called by clustering jointly on the $\Delta$Ct values of \emph{KIR3DL1} and \emph{KIR3DS1}.
%The imputed \emph{KIR3DL1/S1} copy numbers from SNP data were predicted from Log R Ratio and B Allele Frequency with a 1.3\% misclassification error rate
%in leave-one-out cross-validation in the \emph{KIR3DL1/S1} region using a nearest neighbour classifier trained on 747 case and 727 control samples analysed
%both with qPCR and on Immunochip, a custom SNP array.
The association of \emph{KIR3DL1/S1} copy number with T1D was tested in all samples and conditional on the corresponding HLA-Bw4 ligand subsets.
The interaction effect between \emph{KIR3DL1/S1} and HLA-Bw4 was also tested with a case-only analysis.
We can confirm from our study with a substantially increased sample size, that copy number variation in \emph{KIR3DL1/S1} is not an important modifier of T1D risk (0.9 < OR < 1.1 at 95\% confidence for the common genotype), either alone or dependent on HLA-Bw4 epitope, and that furthermore, there is no evidence of statistical interaction effect between HLA-Bw4 and \emph{KIR3DL1/S1} copy number in T1D.  
\emph{( 340 words )}
%This implies that NK are unlikely to be important players in the aetiology of T1D.
%This presents evidence against the role of NK cells in the aetiology of T1D.
%The association is not caused by the interaction with KIR
%These results suggest that HLA Class I susceptibility alleles do not mediate risk to T1D through interplay with KIR3DL1/S1 which would imply that NK cells are not causal of T1D
%These results suggest that the interplay between KIR and HLA may not be an signifcant risk factor in aetology of T1D.
%This strengthens the hypothesis that NK cells are not involved in aetiology of T1D.
}


\doublespacing
%\onehalfspacing

\section{Introduction}
\input{introduction}

\section{Methods}
\input{methods}

\section{Results}
\input{results}

\section{Discussion}
\input{discussion}

\section{Acknowledgements}

This work was funded by the Juvenile Diabetes Research Foundation International, the Wellcome Trust and the National Institute for Health Research (NIHR) Cambridge Biomedical Research Centre.
The research leading to these results has received funding from the European Union's 7th Framework Programme (FP7/2007-2013) under grant agreement no.241447 (NAIMIT).
We thank David Dunger, Barry Widmer, and the British Society for Paediatric Endocrinology and Diabetes for the TID case collection.  
We would like to thank the UK Medical Research Council and Wellcome Trust for funding the collection of DNA for the British 1958 Birth Cohort (MRC grant G0000934, WT grant 068545/Z/02).
DNA control samples were prepared and provided by S. Ring, R. Jones, M. Pembrey, W. McArdle, D. Strachan and P. Burton. 
NP is supported by the Medical Research Council Doctoral Training Grant (C006148).
CW is funded by the Wellcome Trust (089989).
The Diabetes and Inflammation Laboratory is funded by the JDRF, the Wellcome Trust and the National Institute for Health Research (NIHR) Cambridge Biomedical Research Centre.
The Cambridge Institute for Medical Research (CIMR) is in receipt of a Wellcome Trust Strategic Award (100140).

\singlespacing
\footnotesize


% Bibliography
%\bibliographystyle{abbrvnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{chicago}
%\bibliographystyle{apalike}
%\bibliographystyle{naturemag}
%\bibliographystyle{plos2009}

% 32 references
\bibliographystyle{genomeresearch}
\bibliography{paper}



\clearpage


\section*{Figures}
\input{figures}
\FloatBarrier

\clearpage

%\section*{Tables}
%\input{tables}
\FloatBarrier



%\chapter{\label{flowsorting}{Flow Cytometry Cell Sorting}}
%\input{flowsorting}





\end{document}
